---
title: "Second wave of COVID-19 in Switzerland"
author: "Christian L. Althaus"
date: "`r date()`"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    df-print: paged
    embed-resources: true
  pdf:
    toc: true
  docx:
    toc: true
---

```{r}
#| output: false
# Load libraries
library(here)
library(tidyverse)
library(lubridate)
library(unibeCols)
```

## Introduction

The aim of this exercises is to visualize the increase of laboratory-confirmed cases of COVID-19 in Switzerland during autumn 2020.

## Data

We read and process the data of laboratory-confirmed cases of COVID-19 in Switzerland as reported by the [Federal Office of Public Health](https://www.covid19.admin.ch).

```{r}
#| output: false
# Read data
cases <- read_csv(here("data/raw/COVID19Cases_geoRegion.csv"))
```

```{r}
# Process data
region <- "CH"
window_start <- ymd("2020-09-01")
window_end <- ymd("2020-11-30")
cases <- cases |>
  filter(geoRegion == region & datum >= window_start & datum <= window_end) |>
  select(datum, entries)
cases
```

## Results

@fig-cases shows the number of laboratory-confirmed cases of COVID-19 in Switzerland from `r window_start` to `r window_end`.

```{r}
#| label: fig-cases
#| fig-cap: "Laboratory-confirmed cases of COVID-19 in Switzerland in 2020."
#| fig-width: 8
#| fig-height: 4
ggplot(cases, aes(x = datum, y = entries)) +
  geom_bar(stat = "identity", fill = unibeSaphireS()[1]) +
  labs(x = "", y = "Laboratory-confirmed cases", ) +
  theme_minimal()
```

# Conclusions

The number of laboratory-confirmed cases of COVID-19 in Switzerland rapidly increased during October 2020 and subsequently declined.
