---
title: "Basic Statistics Exercises Part 2"
author: "Guy Schnidrig, Beatriz Vidondo"
toc: true
toc-depth: 4
toc-location: left
format: 
  html:
    self-contained: true
    code-link: true
    code-fold: true
    code-summary: "Answer"
number-sections: true
editor: source
theme:
  light: flatly
  dark: darkly
---

These are the exercises for [Basic Statistics and Projects in R](https://zuw.me/kurse/dt.php?kid=4474) course of the [Public Health Sciences Course Program](https://www.medizin.unibe.ch/studies/study_programs/phs_course_program) at the [University of Bern](https://www.unibe.ch/). 

Ask [questions](https://app.sli.do/event/xpuz3U7hL5xXPjtRVmwfw7). 

```{r setup, include=FALSE, echo=F, warning=FALSE, message=FALSE}
# Set language
Sys.setenv(LANG = "en")

# Clear memory
rm(list=ls())
gc()

# Load libraries
library_names <- c("tidyverse", "knitr", "downlit", "ggcorrplot","medicaldata", "rstatix", "datarium","PairedData")

lapply(library_names, function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x)
    library(x, character.only = TRUE)
  }
})

# Set non-scientific notation
options(scipen = 999)
```

## Non-Normality

Load the **cytomegalovirus** data from the package medicaldata^[2](https://cran.r-project.org/web/packages/medicaldata/index.html)^.\
Here you can find the full description^[3](https://www.causeweb.org/tshs/datasets/Cytomegalovirus%20Dataset%20Introduction.pdf)^ of the study, as well as the code book^[4](https://www.causeweb.org/tshs/datasets/Cytomegalovirus%20Data%20Dictionary.pdf)^.

This data set contains 64 consecutive patients who underwent T-cell replete, matched sibling donor reduced-intensity conditioning allogeneic hematopoietic stem cell transplant. The primary risk factor of interest was the number of activating killer immunoglobulin-like receptors (aKIRs: 1-4 vs. 5-6). The primary outcome is **presence** of and **time to cytomegalovirus reactivation**.

```{r, warning=FALSE}
#| code-fold: false
cytomegalovirus <- medicaldata::cytomegalovirus
```

Display the first 5 rows of the data set.

```{r, warning=FALSE}

```

Plot a histogram of the time cytomegalovirus reactivation. Use `ggplot2` and set 20 bins (or use `geom_density`). What kind of distribution do you observe?

```{r}

```

Generate a quantile-quantile (QQ) plot using `geom_qq` and `geom_qq_line` to assess its alignment with the theoretical line.

```{r}

```

We can also test if the distribution is non normal. Use `shapiro_test` to test the null hypothesis that the data came from a normal distribution. Can you reject the null hypothesis?

```{r}

```

Try using `shapiro_test` for multiple variables and save the results in a `tibble`.

```{r}

```

Which type of transformation could you use to make this variable (skew left) look more 'Normal'? Visualize the distribution of transformed variable and and re-test for Normality.

```{r}


```

The Kolmogorov-Smirnov test is a Normality test. Apply this test using function `ks.test` to variables **CD34.dose** and **CD3.dose**. What is the null hypothesis of `ks.test` and can you reject it?

```{r}

```

Calculate if the mean **time to cytomegalovirus** reactivation differs between the sexes. Use the `wilcox_test`.

```{r}

```

Use `geom_boxplot` and `fill` within `ggplot` to visualize your previous result.

```{r}

```

When comparing more than two non-Normal (independent) groups, you can use R function `pairwise_wilcox_test`. In the example below, we create data of students scores in 3 groups A, B and C. Which correction method should you apply? Interpret your results.

```{r}
#| code-fold: false
student_scores <- data.frame(
  Group = rep(c("A", "B", "C"), each = 5),
  Score1 = c(23, 25, 21, 20, 22, 18, 19, 24, 27, 26, 30, 28, 32, 29, 27),
  Score2 = c(19, 18, 17, 16, 13, 20, 15, 22, 25, 24, 26, 4, 30, 22, 13))

student_scores %>%
  kable()
```

```{r}

```

## Correlation

Create a correlation matrix of all numerical measurements in **cytomegalovirus** data. Use `cor`, specify the `spearman` method and only include complete observations. Are there any variables that exhibit a higher degree of correlation with each other?

```{r}

```

Plot the lower half of your correlation matrix with `ggcorplot`.

```{r}


```

Select two variables with a correlation above 0.5 and plot them in a scatter plot (Use `ggplot`). What do you observe?

```{r}


```

Remove the outlier, run the correlation test again and choose the according method. Can you explain what has changed?

```{r}


```

## Dependent Data

Load the data **HorseBeginners** with Actual and Imaginary performances in equitation. Different riders reported how they performed in reality (**Actual**) and according to their own perception (**Imaginary**). In which format is this data (wide or long format)?

```{r}


```

Plot a Bland-Altman plot (`paired.plotBA`) and interpret it, Can you identify any outliers? on which subject?

```{r}

```


Modify (pivot) the format of the dataset from wide into long format. For this you can use functions `pivot_longer` or `gather`and will need to create new variables (Performance and Type). Additionally, convert the categorical variables (Suject and Type) into factors (function `convert_as_factor`). How many times does every subject appear in the data now?

```{r}
HorseBeginnersL <- HorseBeginners %>%
  gather(key = "Type", value = "Performance", Actual,Imaginary) %>%
  convert_as_factor(Subject, Type)

head(HorseBeginnersL, 5) %>%
  kable()
```



Plot a profile plot (box plots of Performance by Type and lines connecting every subject). Can you identify the outlier at S1? Hint: use `geom_boxplot`, `geom_point`, and `geom_line`. 

```{r}

```

Filter out the outlier (for the sake of the exercise).

```{r}

```


Calculate the variability (SumOfSquares) between and within Subjects using the `aov` command. (Hint: Performance ~ Subject)

```{r}

```

In this case, the Subject variable is interpreted as the Between variable. How many groups/clusters/subjects of data do you have?


How much is the total variability (SS-total)?

```{r}

```

Calculate the ICC (for Subject level) and interpret it.

```{r}

```


Advanced exercise: using package **rstatix** set up a [repeated measures Anova](https://www.datanovia.com/en/lessons/repeated) with the **HorseBeginners** and interpret it. Note that this calculation involves both Subject and Type.

```{r}


```

Explore other functions presented in the tutorial like `identify_outliers` and `shapiro_test`. 

Does the dataset HorseBeginnersL really have an outlier? 

```{r}

```



